---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# jlmerclusterperm

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/yjunechoe/jlmerclusterperm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yjunechoe/jlmerclusterperm/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Julia [GLM.jl](https://github.com/JuliaStats/GLM.jl) and [MixedModels.jl](https://github.com/JuliaStats/MixedModels.jl) implementation of bootstrapped cluster-based permutation analysis for time series data, powered by [`JuliaConnectoR`](https://github.com/stefan-m-lenz/JuliaConnectoR)

## Installation

You can install the development version of jlmerclusterperm from [GitHub](https://github.com/yjunechoe/jlmerclusterperm) with:

``` r
# install.packages("remotes")
remotes::install_github("yjunechoe/jlmerclusterperm")
```

Using `jlmerclusterperm` requires a prior installation of Julia, which can be downloaded from either the [official website](https://julialang.org/) or using the command line utility [juliaup](https://github.com/JuliaLang/juliaup). Julia version >=1.8 is required and >=1.9 is preferred for its substantial compiler/runtime improvements.

Before using functions from `jlmerclusterperm`, an initial setup step is required via calling `jlmerclusterperm_setup()`. This will launch Julia with some number of threads (controlled via `options("jlmerclusterperm.nthreads")`) and install necessary package dependencies (this only happens once and should take about 15-30 minutes).

Subsequent calls to `jlmerclusterperm_setup()` incur a small overhead of around 30 seconds and there will be slight delays for first-time function calls due to Julia's [just-in-time compilation](https://docs.julialang.org/en/v1/). Afterwards you can enjoy blazingly-fast functions from the package.

```{r setup}
# Both lines must be run
library(jlmerclusterperm)
system.time(jlmerclusterperm_setup(verbose = FALSE))
```

## Example walkthrough

WIP...

## Notes on R-Julia interoperability

All `jlmerclusterperm` functions collect Julia objects as R objects, except `jlmer()` and `to_jlmer()` which return GLM.jl or MixedModels.jl fitted model objects.

```{r to-jlmer}
jmod <- to_jlmer(Reaction ~ Days + (Days | Subject), lme4::sleepstudy)
jmod # try `glance()` and `tidy()`
```

Model output should be comparable to that of `{lme4}` models:

```{r lmer-comparison, warning=FALSE, message=FALSE}
library(lme4)
rmod <- lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
summary(rmod)
```

Use functions from [`JuliaConnectoR`](https://github.com/stefan-m-lenz/JuliaConnectoR) to interact with these (pointers to) Julia objects:

```{r julia-interop}
library(JuliaConnectoR)
juliaLet("x.rePCA", x = jmod)
juliaCall("issingular", jmod)
```

You can also call functions from other Julia packages that you already have installed. For example, [Effects.jl](https://github.com/beacon-biosignals/Effects.jl) for its [`marginaleffects`](https://github.com/vincentarelbundock/marginaleffects)/[`emmeans`](https://github.com/rvlenth/emmeans)-like features:

```{r effects-jl}
juliaEval("using Effects")
juliaLet("effects(x.namedelements, y)", x = list(Days = unique(sleepstudy$Days)), y = jmod)
```

The following packages are loaded into the Julia environment via `jlmerclusterperm_setup()`:

```{r proj-toml}
cat(readLines(system.file(package = "jlmerclusterperm", "julia", "Project.toml")), sep = "\n")
```

If you wish to use other Julia packages not listed here like Effects.jl, it is recommend to install them in the global library in a fresh Julia session as opposed to `Pkg.add()`-ing using the `JuliaConnectoR` interface, as that may pollute the `jlmerclusterperm` project environment (if this happens, just re-install the package).
