---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "70%"
)
```

# jlmerclusterperm

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Julia [MixedModels.jl](https://github.com/JuliaStats/MixedModels.jl) implementation of cluster-based permutation test for time series data, powered by [{JuliaConnectoR}](https://github.com/stefan-m-lenz/JuliaConnectoR)

## Installation

You can install the development version of jlmerclusterperm from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("yjunechoe/jlmerclusterperm")
```

## Setup

```{r example}
library(jlmerclusterperm)
system.time({jlmerclusterperm_setup()})
```

## Basic example

Example data:

```{r ex-data}
chickweight_df <- as.data.frame(ChickWeight)
fm <- weight ~ 1 + Diet + (1 | Chick)
coplot(
  formula = weight ~ Time | Diet,
  data = chickweight_df,
  panel = function(x, y, ...) {
    points(x, y, ...)
    abline(lm(y ~ x), col = "steelblue", lwd = 3)
  }
)
```

Run Julia mixed model using lme4 syntax, using the convenience form `to_jlmer()`:

```{r lmer-comparison}
library(lme4)
summary(lmer(fm, chickweight_df))

to_jlmer(fm, chickweight_df)
```

Run Julia mixed model directly in `jlmer()` after explicitly reformulating with `jlmer_model_matrix()`:

```{r jlmer-mm-fit}
mm <- jlmer_model_matrix(fm, chickweight_df, cols_keep = "Time")
jlmer(mm$julia_formula, mm$data)
```

Fit jlmer at each timepoint and get a z-matrix back with `jlmer_by_time`:

```{r ex-jlmer-by-time}
by_time <- jlmer_by_time(mm$julia_formula, mm$data, time = "Time")
by_time
```

Find largest cluster in the observed data:

```{r}
empirical_clusters <- largest_clusters(by_time, threshold = 1.5)
empirical_clusters
```

We can think of the data as having a within-participant design, since each chick is assigned one diet. `clusterpermute()` conducts a cluster-based permutation test by randomly shuffling the assignment of diet between chicks to generate a null distribution of the largest sum-z cluster by predictor. 

```{r}
system.time({
  null_dist_clusters <- clusterpermute(
    julia_formula = mm$julia_formula, data = mm$data, time = "Time",
    nsim = 1000, participant_col = "Chick", threshold = 1.5,
    binned = FALSE # For non-aggregated data
  )
})
```

Test empirical clusters against the simulated null:

```{r}
lapply(c("Diet2", "Diet3", "Diet4"), function(predictor) {
  empirical <- empirical_clusters[[predictor]]$sum_z
  null_dist <- null_dist_clusters[[predictor]]
  mean(abs(null_dist) > empirical)
})
```

## Formula utilities

```{r}
jlmer_model_matrix(mpg ~ wt * qsec + (1 + wt | vs), head(mtcars))
jlmer_model_matrix(mpg ~ wt * qsec + (1 + wt || vs), head(mtcars))
jlmer_model_matrix(mpg ~ wt * qsec + (1 + wt | vs), head(mtcars), drop_terms = "wt__qsec")
jlmer_model_matrix(mpg ~ wt * qsec + (1 + wt | vs), head(mtcars), cols_keep = TRUE)
```


