---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# jlmerclusterperm

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Julia [GLM.jl](https://github.com/JuliaStats/GLM.jl) and [MixedModels.jl](https://github.com/JuliaStats/MixedModels.jl) implementation of bootstrapped cluster-based permutation analysis for time series data, powered by [{JuliaConnectoR}](https://github.com/stefan-m-lenz/JuliaConnectoR)

## Installation

You can install the development version of jlmerclusterperm from [GitHub](https://github.com/yjunechoe/jlmerclusterperm) with:

``` r
# install.packages("remotes")
remotes::install_github("yjunechoe/jlmerclusterperm")
```

Using `jlmerclusterperm` requires a prior installation of Julia, which can be downloaded from either the [official website](https://julialang.org/) or via the command line utility [juliaup](https://github.com/JuliaLang/juliaup). Julia version >=1.8 is required and >=1.9 is preferred for its substantial compiler/runtime improvements.

Before using functions from `jlmerclusterperm`, an initial setup step is required via calling `jlmerclusterperm_setup()`. This will launch Julia with some number of threads (controlled via `options("jlmerclusterperm.nthreads")`) and install necessary Julia package dependencies (this only happens once and should take about 15-30 minutes). Every call to `jlmerclusterperm_setup()` after the initial setup incurs a small overhead (around 30 seconds to 1 minute) after which you will get access to multi-threaded, blazingly-fast functions in the package.

```{r setup}
# Both lines must be run for a new R session
library(jlmerclusterperm)
jlmerclusterperm_setup()
```

## Example walkthrough

WIP...

## Notes on R-Julia interoperability

All `jlmerclusterperm` functions collect Julia objects as R objects, except `jlmer` and `to_jlmer` which return either GLM.jl or MixedModels.jl fitted model objects.

```{r jmod}
jmod <- to_jlmer(Reaction ~ Days + (Days | Subject), sleepstudy)
jmod
```

You can use functions from `JuliaConnectoR` to interact with these (pointers to) Julia objects:

```{r}
library(JuliaConnectoR)
juliaLet("x.rePCA", x = jmod)

# For {marginaleffects} / {emmeans} features
juliaEval("using Effects")
# For converting to R {lme4} objects
juliaEval("using JellyMe4")
```

