---
title: "Asynchronous CPA"
description: |
  Using {future} to run CPA asynchronously in a background process
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette showcases using `{future}` to run CPA asynchronously in a background process.

See more tutorials and vignettes on the [Articles](https://yjunechoe.github.io/jlmerclusterperm/articles) page.

### Setup

```{r setup}
library(jlmerclusterperm)
jlmerclusterperm_setup(verbose = FALSE)
```

Minimal example data and CPA specification:

```{r MWE-setup, message = FALSE, warning = FALSE}
chickweights <- ChickWeight
chickweights$Time <- as.integer(factor(chickweights$Time))
chickweights_spec <- make_jlmer_spec(
  formula = weight ~ 1 + Diet,
  data = chickweights,
  subject = "Chick", time = "Time"
)
```

Example CPA (1000 simulations):

```{r ex-CPA}
set_rng_state(123L)
CPA <- clusterpermute(
  chickweights_spec,
  threshold = 2.5,
  nsim = 1000,
  progress = FALSE
)
```

<div class="callout" style="font-size:0.9em">
The details of the actual CPA is skipped because that's not the focus of this vignette, but this is the same example from the [readme](https://github.com/yjunechoe/jlmerclusterperm/blob/main/README.md) just with more `nsim`.
</div>

## Basic idea

The basic idea behind this asynchronous CPA strategy is to start a background R process whose sole job is to send the instructions for CPA over to Julia. It's **asynchronous** because running the CPA this way **will not block your interactive R session**. Since most of the work is being done in Julia anyways, this parallelization has virtually no impact on the performance of evaluating R code.

The `{future}` package allows such asynchronous evaluation of R code. It's a big package implementing a complex topic. You can read more about the project on the [futureverse](https://www.futureverse.org/) but we just show the bare minimum here.

For more advanced users, note that we're use parallelization _only_ for its asynchronous properties (non-blocking evaluation of R code in a background process). It is not recommended to start _multiple R processes_ with each running a CPA because the Julia session is shared abd already multithreaded (if you can spare more cores, set that in `options("jlmerclusterperm.nthreads")` before calling `jlmerclusterperm_setup()`).

Later updates to `jlmerclusterperm` may wrap this workflow in a more principled way, but for now this vignette serves as a minimally working example for asynchronously running a CPA.

## Walkthrough

### Setup for async CPA

Three things are in order for this workflow:

1. Load the `{future}` package
2. Initialize the `multisession` future
3. Grab options from the `jlmerclusterperm` package environment

```{r future-setup}
library(future)
plan(multisession)
pkgopts <- as.list(jlmerclusterperm:::.jlmerclusterperm)
```

<div class="callout" style="font-size:0.9em">
Please treat the `.jlmerclusterperm` internal variable as a read-only object - it's unexported and not meant to be manipulated.
</div>


### Creating the future object

We start by creating a special object of class `<Future>` using `future::future()`. We can simply pass the `clusterpermute()` code to the first argument of `future()`, but with a slight modification to ensure that the background process connects to the same Julia session.

The template is as follows. First, you make the `pkgopts` object (defined above) available for the future (via the `globals` argument). Then, inside the future expression ensure that the background process shares the same Julia session (via `list2env(...)`).

```{r eval = FALSE}
# Not run
future(
  {
    list2env(pkgopts, jlmerclusterperm:::.jlmerclusterperm)
    ## Your CPA code below ##
  },
  globals = structure(TRUE, add = "pkgopts")
)
```

A future object replicating the 1000-simulation CPA from above will look like the following:

```{r future-cpa}
f <- future(
  {
    list2env(as.list(pkgopts), jlmerclusterperm:::.jlmerclusterperm)
    ## Your CPA code below ##
    set_rng_state(123L)
    clusterpermute(
      chickweights_spec,
      threshold = 2.5,
      nsim = 1000
    )
  },
  globals = structure(TRUE, add = "pkgopts")
)
```

Once the future is created, it **immediately** starts executing the code in a background process. It is important that you **do not evaluate the future object (`f`) directly**. Instead, you should query it with `future::resolved()` - this simply tells you whether the background evaluation has completed or not:

```{r resolved}
resolved(f)
```

Here, we use a `while` loop to show that the background CPA is non-blocking. We are evaluating R code in the interactive session _simultaneously_ as the CPA is running:

```{r resolved-loop, results='hold'}
i <- 0
while (!resolved(f)) {
  Sys.sleep(0.5)
  i <<- i + 1
  cat(sprintf("Elapsed: %.01fs", i / 2), "\n")
}
```

We reach the end of the while loop at `r i/2` seconds, which is approximately how long the CPA took to complete (we confirm this in the next section). At this point the future has completed and its result is available for collection.

```{r resolved}
```

<div class="callout" style="font-size:0.9em">
Although the R code here was just a crude progress alert using `Sys.sleep()`, you can freely evaluate _any_ R code **except other functions that call Julia**. You can make plots, clean data, write up analyses, etc., but just don't run another CPA while one is already running because the Julia process is shared.
</div>

### Collecting the results

We can collect the output of the background process with `future::value()`. But first, we double check to make sure that the background process has indeed finished evaluating:

```{r resolved}
```

Once that's confirmed, we can use `value()` to collect the results from the future and assign it to a variable for further inspection:

```{r CPA_async}
CPA_async <- value(f)
```

Note how `value()` also prints the messages encountered while executing the CPA code. We see this because we did not specify `progress = FALSE` in the `clusterpermute()` call passed to the future. For our purposes, just note that adding up the times from the messages is similar to what we saw from the while loop (`r i/2` seconds).

The output of the asynchronous CPA (`CPA_async`) is identical to the initial CPA that we ran at the beginning (`CPA`) because both were ran with the default seed of 1 and the RNG counter value of `123L`:

```{r CPAs-identical}
identical(CPA_async, CPA)
```

<div class="callout" style="font-size:0.9em">
A separate vignette covers the [Julia RNG](https://yjunechoe.github.io/jlmerclusterperm/articles/reproducibility.html).
</div>

In practice, you may want to use a random seed (via `set_rng_seed()`) for the background CPA.
