---
title: "Julia interface"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(jlmerclusterperm)
jlmerclusterperm_setup(verbose = FALSE)
julia_progress(show = FALSE)
```

## Notes on R-Julia interoperability

All `jlmerclusterperm` functions collect Julia objects as R objects, except `jlmer()` and `to_jlmer()` which return GLM.jl or MixedModels.jl fitted model objects.

```{r to-jlmer}
jmod <- to_jlmer(Reaction ~ Days + (Days | Subject), lme4::sleepstudy, REML = TRUE)
jmod
```

Model output should be comparable to that of `{lme4}` models:

```{r lmer-comparison, warning=FALSE, message=FALSE}
library(lme4)
rmod <- lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
summary(rmod)
```

You can use functions from [`JuliaConnectoR`](https://github.com/stefan-m-lenz/JuliaConnectoR) to interact with these (pointers to) Julia objects:

```{r julia-interop}
library(JuliaConnectoR)
juliaLet("x.rePCA", x = jmod)
juliaCall("issingular", jmod)
```

You can also call functions from other Julia packages that you already have installed. For example, [Effects.jl](https://github.com/beacon-biosignals/Effects.jl) for its [`marginaleffects`](https://github.com/vincentarelbundock/marginaleffects)/[`emmeans`](https://github.com/rvlenth/emmeans)-like features:

```{r effects-jl}
juliaEval("using Effects")
juliaLet("effects(x.namedelements, y)", x = list(Days = 2:5), y = jmod)
```

The following packages are loaded into the Julia environment via `jlmerclusterperm_setup()`:

```{r proj-toml}
cat(readLines(system.file(package = "jlmerclusterperm", "julia", "Project.toml")), sep = "\n")
```

If you wish to use other Julia packages not listed here like Effects.jl, it is recommend to install them in the global library in a fresh Julia session as opposed to `Pkg.add()`-ing using the [`JuliaConnectoR`](https://github.com/stefan-m-lenz/JuliaConnectoR) interface, as that may pollute the `jlmerclusterperm` project environment (if this happens, just re-install the package).
