---
title: "Julia interface"
description: |
  Interacting with Julia regression models from R using {JuliaConnectoR}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

```{r setup}
library(jlmerclusterperm)
jlmerclusterperm_setup(verbose = FALSE)
julia_progress(show = FALSE)
```

## Interacting with Julia objects

All `jlmerclusterperm` functions collect Julia objects as R objects, except `jlmer()` and `to_jlmer()` which return GLM.jl or MixedModels.jl fitted model objects.

```{r to-jlmer}
jmod <- to_jlmer(Reaction ~ Days + (Days | Subject), lme4::sleepstudy)
jmod
#> <Julia object of type LinearMixedModel{Float64}>
#> Variance components:
#>             Column    Variance Std.Dev.   Corr.
#> Subject  (Intercept)  565.51067 23.78047
#>          Days          32.68212  5.71683 +0.08
#> Residual              654.94145 25.59182
#> ──────────────────────────────────────────────────
#>                 Coef.  Std. Error      z  Pr(>|z|)
#> ──────────────────────────────────────────────────
#> (Intercept)  251.405      6.63226  37.91    <1e-99
#> Days          10.4673     1.50224   6.97    <1e-11
#> ──────────────────────────────────────────────────
```

You can use functions from [`JuliaConnectoR`](https://github.com/stefan-m-lenz/JuliaConnectoR) to interact with these (pointers to) Julia objects:

```{r julia-interop}
library(JuliaConnectoR)
juliaLet("x.rePCA", x = jmod)
#> <Julia object of type NamedTuple{(:Subject,), Tuple{Vector{Float64}}}>
#> (Subject = [0.5406660682947617, 1.0],)
juliaCall("issingular", jmod)
#> [1] FALSE
```

You can also call functions from other Julia packages that you already have installed. For example, [Effects.jl](https://github.com/beacon-biosignals/Effects.jl) for its [`marginaleffects`](https://github.com/vincentarelbundock/marginaleffects)/[`emmeans`](https://github.com/rvlenth/emmeans)-like features:

```{r effects-jl}
juliaEval("using Effects")
juliaLet("effects(x.namedelements, y)", x = list(Days = 2:5), y = jmod)
#> <Julia object of type DataFrame>
#> 4×5 DataFrame
#>  Row │ Days   Reaction  err      lower    upper
#>      │ Int64  Float64   Float64  Float64  Float64
#> ─────┼────────────────────────────────────────────
#>    1 │     2   272.34   7.09427  265.245  279.434
#>    2 │     3   282.807  7.70544  275.102  290.512
#>    3 │     4   293.274  8.55557  284.719  301.83
#>    4 │     5   303.742  9.58128  294.16   313.323
```

Note that `jlmer()` and `to_lmer()` are just conveniences for sanity checking the steps of a cluster-based permutation test, and thus should not be used as a general interface to fitting regression models in Julia for more serious analyses.

## Julia global options

All functions involved in the cluster-based permutation analysis print progress bars that update in real time in the console. The function `julia_progress()` controls whether to `show` the progress bar and the `width` of the printed progress bar.

```{r julia_progress}
# Set Julia progress options and save old state
old_progress_opts <- julia_progress(show = FALSE, width = 30)
old_progress_opts
#> $show
#> [1] TRUE
#>
#> $width
#> [1] 50
print( julia_progress() )
#> $show
#> [1] FALSE
#>
#> $width
#> [1] 30

# Restored old state
julia_progress(old_progress_opts)
identical(old_progress_opts, julia_progress())
#> [1] TRUE
```

## The Julia environment

The following packages are loaded into the Julia environment via `jlmerclusterperm_setup()`:

```{r proj-toml, eval = TRUE}
cat(readLines(system.file("julia/Project.toml", package = "jlmerclusterperm")), sep = "\n")
```

If you wish to use other Julia packages not listed here like Effects.jl, it is recommend to install them in the global library in a fresh Julia session as opposed to `Pkg.add()`-ing using the [`JuliaConnectoR`](https://github.com/stefan-m-lenz/JuliaConnectoR) interface, as that may pollute the `jlmerclusterperm` project environment (if this happens, just re-install the package).
