---
title: "Tidying output"
description: |
  Collecting {jlmerclusterperm} objects as tidy data frames
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette showcases `tidy()` and other ways of collecting `{jlmerclusterperm}` objects as [tidy data](https://r4ds.had.co.nz/tidy-data.html).


### Setup

```{r setup}
library(jlmerclusterperm)
jlmerclusterperm_setup(verbose = FALSE)
julia_progress(show = FALSE)
```

Minimal example data:

```{r MWE-setup, message = FALSE, warning = FALSE}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
chickweights <- as_tibble(ChickWeight) %>%
  mutate(diet = forcats::fct_collapse(Diet, A = c(1, 3), B = c(2, 4))) %>%
  filter(Time %in% 0:20)
ggplot(chickweights, aes(Time, weight, color = diet)) +
  stat_summary(fun.args = list(mult = 1.96)) +
  stat_summary(geom = "line") +
  theme_minimal(base_size = 18)
```

The jlmer specification object used for this showcase:

```{r}
jlmer_spec <- make_jlmer_spec(
  formula = weight ~ diet + (1 | Chick), data = chickweights,
  subject = "Chick", time = "Time"
)
jlmer_spec
```


## A tour of `tidy()` methods

All non-`<data.frame>` objects returned by `jlmerclusterperm` functions have a `tidy()` method to return a data frame representation of the object.

### `<timewise_statistics>` objects

Functions `compute_timewise_statistics()` and `permute_timewise_statistics()` return an object of class `<timewise_statistics>`, which is an array of statistics. Note that there isn't a stylized `print()` method for this class.

Instead, it is recommended to use `tidy()` to inspect the results as a data frame where each statistic is represented as a row.

In the case of `compute_timewise_statistics()`, the output is a matrix (2D array) of predictor and time. Each dimension becomes a column in the `tidy()` data:

```{r tidy-timewise_statistics}
empirical_statistics <- compute_timewise_statistics(jlmer_spec)
class(empirical_statistics)
dim(empirical_statistics) # predictor x time
tidy(empirical_statistics)
```

In the case of `compute_timewise_statistics()`, the output is a 3D array of simulation, time, and predictor. Again, each dimension becomes a column in the `tidy()` data:

```{r tidy-compute_timewise_statistics}
null_statistics <- permute_timewise_statistics(jlmer_spec, nsim = 1000)
class(null_statistics)
dim(null_statistics) # simulation x time x predictor
tidy(null_statistics)
```

### `<empirical_clusters>` objects

The `<empirical_clusters>` object returned by `extract_empirical_clusters()` does have a print method defined, which makes it easier on the eye:

```{r empirical_clusters}
empirical_clusters <- extract_empirical_clusters(empirical_statistics, threshold = 1)
class(empirical_clusters)
empirical_clusters
```

The `tidy()` data frame representation also contains information like `length`, which is the number of time points that the cluster spans. The `id` variable uniquely identifies a cluster within each predictor.

```{r tidy-empirical_clusters}
tidy(empirical_clusters)
```

### `<null_cluster_dists>` objects

The `<null_cluster_dists>` object returned by `extract_null_cluster_dists()` also has a print method defined, which displays summary statistics:

```{r null_cluster_dists}
null_cluster_dists <- extract_null_cluster_dists(null_statistics, threshold = 1)
class(null_cluster_dists)
null_cluster_dists
```

The null distribution is constructed from the largest cluster-mass statistic observed in each of the bootstrapped permutations of the original data. In other words, each simulation contributes one sample to the null.

In the `tidy()` representation, each simulation of a predictor represents a row. The `sum_statistic` column represents the size of the _largest_ cluster from each simulation. When no clusters are found, the column is zero and values for `start`, `end`, and `length` are `NA`.

```{r tidy-null_cluster_dists}
tidy(null_cluster_dists)
```

The output of `tidy(null_statistics)` and `tidy(null_cluster_dists)` form a relational data. For example, you can use `left_join()` to find the time wise statistics of the permuted data with the most extreme cluster-mass statistic:

```{r largest_permuted_cluster}
largest_permuted_cluster <- tidy(null_cluster_dists) %>%
  slice(which.max(abs(sum_statistic))) %>%
  left_join(tidy(null_statistics), by = c("predictor", "sim"))
largest_permuted_cluster
```

Note that the `sum_statistic` is simply the sum of the `statistic` values over the span of the largest cluster:

```{r sum_statistic-by-hand}
largest_permuted_cluster %>%
  filter(between(time, start, end)) %>%
  summarize(sum_statistic = unique(sum_statistic), sum(statistic))
```

### In other contexts

The output of `calculate_clusters_pvalues()` is just an `<empirical_clusters>` object that has been augmented with p-values:

```{r empirical_clusters_tested}
tested_clusters <- calculate_clusters_pvalues(empirical_clusters, null_cluster_dists)
tested_clusters
```

When available, p-values are returned as a column in the `tidy()` data:

```{r tidy-empirical_clusters_tested}
tidy(tested_clusters)
```

Note that the p-value for this cluster is `r tidy(tested_clusters)$pvalue` because there are `r as.integer(tidy(tested_clusters)$pvalue * 1000)` samples from the null that are as or more extreme than the observed:

```{r pvalue-by-hand}
tidy(null_cluster_dists) %>%
  filter(abs(sum_statistic) >= abs(tidy(tested_clusters)$sum_statistic))
```

In the case of the one-fell-swoop function `clusterpermute()`, the `<null_cluster_dists>` object and the `<empirical_clusters>` object are returned in a list:

```{r clusterpermute}
# Timing the entire process this time
system.time({
  full_test <- clusterpermute(jlmer_spec, threshold = 1, nsim = 1000, progress = FALSE)
})
full_test
```

You can `tidy()` the elements of this list with `lapply()` or `purrr::map()`:

```{r tidy-clusterpermute}
lapply(full_test, tidy)
```


## Tidying Julia model objects

Functions `jlmer()` and `to_jlmer()` return an object of class `<jlmer_mod>`, which is a [`JuliaConnectoR`](https://github.com/stefan-m-lenz/JuliaConnectoR) pointer to the Julia model object:

```{r jlmer_mod}
jlmer_mod <- to_jlmer(weight ~ diet + (1 | Chick), chickweights)
class(jlmer_mod)
jlmer_mod
```

You can use `tidy()` and `glance()` to return information about the Julia model as a data frame:

```{r tidy-glance}
tidy(jlmer_mod)
glance(jlmer_mod)
```

These mimic the behavior of the same methods from `{broom}`/`{broom.mixed}` defined for `(g)lm`/`(g)lmer` models:

```{r lme4, message = FALSE, warning = FALSE}
library(lme4)
library(broom.mixed)
lmer_mod <- lmer(weight ~ diet + (1 | Chick), chickweights, REML = FALSE)
tidy(lmer_mod)
glance(lmer_mod)
```

Note that Julia [MixedModels.jl](https://github.com/JuliaStats/MixedModels.jl) defaults to `REML = FALSE`, whereas it is `TRUE` by default in `{lme4}`.
